# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
act <- read.csv(file = "activity.csv",
                header = T,
                col.names = c("Steps", "Date", "Interval"),
                colClasses = c("numeric", "Date", "numeric"))

```

## What is mean total number of steps taken per day?

Histogram of the total number of steps taken each day with indicated mean value blue and median value red
```{r fig.width=7, fig.height=6, echo=FALSE}
library(ggplot2)

## factoring the data
factors <- levels(factor(act$Date))
sums <- numeric()
for(i in 1:length(factors)) {
    sums[i] <- sum(act[act$Date==factors[i], 1], na.rm=T)
    }
## histogram with indicated mean value blue and median value red
```
```{r fig.width=7, fig.height=6, echo=TRUE}
ggplot(act, aes(x=sums)) +
    geom_histogram(binwidth=1000, colour="black", fill="white") + 
    geom_vline(aes(xintercept=mean(sums, na.rm=T)),
               color="blue", linetype="dashed", size=1) +
    geom_vline(aes(xintercept=median(sums, na.rm=T)),
               color="red", linetype="solid", size=1)
```

**MEAN** and **MEDIAN** total number of steps taken per day
```{r echo=FALSE}
mean <- mean(sums)
median <- median(sums)
```
Mean = `r mean`;
Median = `r median`

## What is the average daily activity pattern?
```{r echo=FALSE}
sumint <- as.numeric()
n <- 1
for(i in act[act$Date=="2012-10-01", 3]){
    sumint[n] <- mean(act[act$Interval==i, 1], na.rm=T)
    n <- n + 1
    }
```
```{r echo=TRUE}
ggplot(act, aes(x=act[act$Date=="2012-10-01", 3], y=sumint)) +
    labs(x = "Interval", y = "Average amount of steps") +
    geom_line()
```

```{r echo=FALSE}
maxintrv <- act[sumint==max(sumint), 3]
```

5-minute interval, taken from average across all the days in the dataset, containing the maximum number of steps is `r maxintrv[1]`.

```{r echo=FALSE}
numna <- length(act[is.na(act$Steps), 1])
```

Total number of missing values in the dataset = `r numna`.

## Imputing missing values
```{r echo=FALSE}
## creates a new dataset
act2 <- act
missingvalues <- as.numeric()

## make a new column with means of 5-minute intervals
for(i in 0:(length(act[, 1])-1)){
    missingvalues[i+1] <- sumint[(i %% length(sumint)) + 1]
    }   

## change NA values with the mean of 5-minute intervals
act2[is.na(act$Steps), 1] <- missingvalues[is.na(act$Steps)]
```

Histogram of the total number of steps taken each day with NA values changed
```{r echo=FALSE}
sums2 <- numeric()
for(i in 1:length(factors)) {
    sums2[i] <- sum(act2[act2$Date==factors[i], 1], na.rm=T)
    }

## histogram with indicated mean and median value (same line)
```
```{r fig.width=7, fig.height=6, echo=TRUE}
ggplot(act2, aes(x=sums2)) +
    geom_histogram(binwidth=1000, colour="black", fill="white") + 
    geom_vline(aes(xintercept=mean(sums2, na.rm=T)),
               color="blue", linetype="dashed", size=1) +
    geom_vline(aes(xintercept=median(sums2, na.rm=T)),
               color="red", linetype="solid", size=1)
```

**MEAN** and **MEDIAN** total number of steps taken per day
```{r echo=FALSE}
mean2 <- mean(sums2)
median2 <- median(sums2)
```
Mean without NA = `r mean2`;
Median without NA = `r median2`

*Do these values differ from the estimates from the first part of the assignment?*
Yes, the values are different. There is significant change in histogram at around 0 and 10000 steps. Both Mean and median has changed.

*What is the impact of imputing missing data on the estimates of the total daily number of steps?*
Mean and median values has changed. 

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=FALSE}
## for plotting two additional libraries are going to be used
library(grid)
library(gridExtra)

act[, 4] <- as.character(length(act[, 2]))
names(act)[4]<-paste("Weekday") 
for(i in 1:length(act[, 2])){
    if(weekdays(as.Date(act[i, 2])) == "Saturday" | 
           weekdays(as.Date(act[i, 2])) == "Sunday"){
        act[i, 4] <- "Weekend"
        } else {
            act[i, 4] <- "Weekday"
            }
    }

avewkend <- as.numeric()
avewkday <- as.numeric()
n <- 1
for(i in act[act$Date=="2012-10-01", 3]){
    avewkend[n] <- mean(act[act$Interval==i & act$Weekday=="Weekend", 1], 
                        na.rm=T)
    avewkday[n] <- mean(act[act$Interval==i & act$Weekday=="Weekday", 1], 
                        na.rm=T)
    n <- n + 1
    }
```
```{r fig.width=7, fig.height=6, echo=TRUE}
p1 <- ggplot(act, aes(x=act[act$Date=="2012-10-01", 3], y=avewkend)) +
    labs(x = "", y="Weekend") +
    geom_line()
p2 <- ggplot(act, aes(x=act[act$Date=="2012-10-01", 3], y=avewkday)) +
    labs(x = "", y="Weekdays") +
    geom_line()
grid.arrange(p1, p2, nrow = 2, left= "Average amount of steps", sub="Interval")
```
